Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving

### 2022-02-25 paper/main.tex
preliminaries polished

\newcommand{\Nset}{{\mathbb{N}_0}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
%\def\Inf{\ensuremath{\mathsf{Inf}}}
%\def\Fin{\ensuremath{\mathsf{Fin}}}
\DeclareMathOperator{\Inf}{\mathsf{Inf}}
\DeclareMathOperator{\Fin}{\mathsf{Fin}}
\newcommand{\minf}{\mathit{inf}}
\def\false{\mathit{false}}
\def\true{\mathit{true}}
\newcommand{\mA}{\mathcal{A}}
\newcommand{\mB}{\mathcal{B}}

@@

The number of acceptance marks can be algorithmically reduced to one
as every TELA can be transformed to an equivalent Büchi automaton
(this can be easily done for example by Spot~\cite{duret.16.atva2}),
but this reduction is paid by dramatic changes of state-space: the
number of states can increase exponentially and some important
structural properties like determinism can be lost. This motivates our
study of a technique reducing the number of acceptance marks without
altering the structure of the automaton.

We present a technique heavily based on \emph{quantified Boolean
  formulas (QBF)}. For a given TELA and parameters $k,n$, it produces
a QBF formula which is satisfiable if and only if there exists an
automaton with the same structure, $n$ acceptance marks, an acceptance
formula in disjunctive normal form with $k$ clauses, and the same set
of accepting runs as the original automaton. The placement of the
marks on transitions and the acceptance formula can be obtained from a
model of the formula. Besides this formula, we describe also the
construction of two simpler formulas such that their satisfiability
implies the existence of an automaton with the same structure, $n$
acceptance marks, and the same set of accepting runs, but not vice
versa.

@@

A tuple $t=(p,a,M',q)\in\delta$ is the \emph{transition} leading from
state $p$ to state $q$ labelled with $a$ and acceptance marks $M'$. We
use $M(t)$ to refer to the set $M'$ of acceptance marks. Further, for
a set $T\subseteq\delta$ of transitions, $M(T)=\bigcup_{t\in T}M(t)$
denotes the set of acceptance marks that appear on transitions in $T$.

A run $\pi$ of $\mA$ over an infinite word
$u = u_{0}u_{1}u_{2}\dots \in \Sigma^{\omega}$ is an infinite sequence
of adjacent transitions
$\pi=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots\in\delta^{\omega}$ where
$q_0=q_I$. Let $\minf(\pi)$ denote the set of transitions that appear
infinitely many times in $\pi$. Further, let $M(\pi)$ denote the set
of acceptance marks that occur in infinitely many transitions of run
$\pi$, i.e., $M(\pi)=M(\minf(\pi))$. Run $\pi$ is \emph{accepting} iff
$M(\pi)$ satisfies the formula $\varphi$, where $M(\pi)$ satisfies
$\Inf m$ iff $m\in M(\pi)$ and $M(\pi)$ satisfies $\Fin m$ iff
$m\not\in M(\pi)$. The \emph{language} of automaton $\mA$ is the set
$L(\mA)=\{u\in \Sigma^{\omega}\mid\textrm{there is an accepting run of
}\mA\textrm{ over }u\}$. Two automata $\mA,\mB$ are \emph{equivalent}
if $L(\mA)=L(\mB)$.
% such that $\mA$ has
% an accepting run over $u$.

@@

We start with the several simple observations. Acceptance of a run
$\pi$ is fully determined by the set $M(\pi)$ of acceptance marks that
appear infinitely often on $\pi$, which is in turn fully determined by
the set $\inf(\pi)$ of transitions appearing infinitely often in
$\pi$.
% \[
%   M(\pi)=\bigcup_{(q_i,u_i,M_i,q_{i+1})\in\minf(\pi)}M_i
% \]
Further, each set of the form $\inf(\pi)$ is a cycle and vice versa.
Hence, our reduction methods is looking for

following simple observation. Definovat $inf(\varphi)$?

>>>>

\newcommand{\Nset}{{\mathbb{N}_0}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
%\def\Inf{\ensuremath{\mathsf{Inf}}}
%\def\Fin{\ensuremath{\mathsf{Fin}}}
\DeclareMathOperator{\Inf}{\mathsf{Inf}}
\DeclareMathOperator{\Fin}{\mathsf{Fin}}
\newcommand{\minf}{\mathit{inf}}
\newcommand{\mks}{\mathit{mks}}
\def\false{\mathit{false}}
\def\true{\mathit{true}}
\newcommand{\mA}{\mathcal{A}}
\newcommand{\mB}{\mathcal{B}}

@@

The number of acceptance marks can be algorithmically reduced to one
as every TELA can be transformed to an equivalent Büchi automaton
(this can be easily done for example by Spot~\cite{duret.16.atva2}),
but this reduction is paid by dramatic changes of state-space: the
number of states can increase exponentially in the number of
acceptance marks and some important structural properties like
determinism can be lost. This motivates our study of a technique
reducing the number of acceptance marks without altering the structure
of the automaton.

We present such a technique heavily based on \emph{quantified Boolean
  formulas (QBF)}. For a given TELA and parameters $k,n$, it produces
a QBF formula which is satisfiable if and only if there exists an
automaton with the same structure, $n$ acceptance marks, an acceptance
formula in disjunctive normal form with $k$ clauses, and the same set
of accepting runs as the original automaton. The placement of the
marks on transitions and the acceptance formula can be obtained from a
model of the formula. Besides this formula, we describe also the
construction of two simpler formulas such that their satisfiability
implies the existence of an automaton with the same structure, $n$
acceptance marks, and the same set of accepting runs, but not vice
versa.

@@

A tuple $t=(p,a,M',q)\in\delta$ is the \emph{transition} leading from
state $p$ to state $q$ labelled with $a$ and acceptance marks $M'$.  A
run $\pi$ of $\mA$ over an infinite word
$u = u_{0}u_{1}u_{2}\dots \in \Sigma^{\omega}$ is an infinite sequence
of adjacent transitions
$\pi=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots\in\delta^{\omega}$ where
$q_0=q_I$. Let $\minf(\pi)$ denote the set of transitions that appear
infinitely many times in $\pi$.
%
We use $\mks(\cdot)$ to denote the set of acceptance
marks of various objects:
\begin{itemize}
\item for a transitions $t=(p,a,M',q)\in\delta$, $\mks(t)$ denotes the set $M'$,
\item for a set $T\subseteq\delta$ of transitions,
  $\mks(T)=\bigcup_{t\in T}\mks(t)$ denotes the set of marks that appear on
  transitions in $T$,
\item for a run $\pi$, $\mks(\pi)=\mks(\minf(\pi))$ denotes the set of
  marks that occur in infinitely many transitions of run $\pi$.
\end{itemize}
%
Run $\pi$ is \emph{accepting} iff $\mks(\pi)$ satisfies the formula
$\varphi$, where a set of marks $M'\subseteq M$ satisfies $\Inf m$ iff
$m\in M'$ and it satisfies $\Fin m$ iff $m\not\in M'$. The
\emph{language} of automaton $\mA$ is the set
$L(\mA)=\{u\in \Sigma^{\omega}\mid\textrm{there is an accepting run of
}\mA\textrm{ over }u\}$. Two automata $\mA,\mB$ are \emph{equivalent}
if $L(\mA)=L(\mB)$.
% such that $\mA$ has
% an accepting run over $u$.

@@

We start with the several simple observations. Acceptance of a run
$\pi$ is fully determined by the set $M(\pi)$ of acceptance marks that
appear infinitely often on $\pi$, which is in turn fully determined by
the set $\minf(\pi)$ of transitions appearing infinitely often in
$\pi$.
% \[
%   M(\pi)=\bigcup_{(q_i,u_i,M_i,q_{i+1})\in\minf(\pi)}M_i
% \]
Further, each set of the form $\minf(\pi)$ is a cycle and vice versa.
Hence, our reduction methods is looking for a new acceptance formula
$\psi$ with less accceptance marks than the original acceptance
formula $\varphi$ and for new placement of the marks on transitions
such that for each cycle $C\subseteq\delta$ it holds that the marks $M(C)$
satisfies $\varphi$ if and only if $M(C)$ satisfies $\psi$.
